---
title: "HW6"
author: "Matthew Huang (PID: A17978767)"
format: pdf
---

## Code

```{r}
analyze_protein_bfactors <- function(pdb_ids, plot_id = pdb_ids[1], 
                                     chain = "A", elety = "CA", show_sse = TRUE) 
  {
  # Inputs:
  # - pdb_ids is the character vector of PDB IDs, like c("4AKE","1AKE","1E4Y")
  # - plot_id is the single PDB ID from pdb_ids that you want to plot
  # - chain is the chain kept with default "A"
  # - elety is the atom type kept with default "CA"
  # - show_sse: shows TRUE/FALSE; if FALSE it will remove the marginal SSE rectangles
  
  # What the function does:
  # For each PDB ID, it reads structure, then trims to selected chain+atom type
  # extracts the B-factors, and then plots B-factors for the requested plot_id.
  
  # Output:
  # Returns a named list of B-factor vectors with one element per input PDB ID.
  
  library(bio3d)
  
  # Initial Checks
  stopifnot(is.character(pdb_ids))
  stopifnot(length(pdb_ids) >= 1)
  stopifnot(plot_id %in% pdb_ids)
  
  b_factors <- list()
  sse_info <- list()
  
  # Loop over any number of input structures to generalizes s1/s2/s3
  for (id in pdb_ids) {
    pdb <- read.pdb(id)
    pdb_trim <- trim.pdb(pdb, chain = chain, elety = elety)
    
    b_factors[[id]] <- pdb_trim$atom$b
    sse_info[[id]] <- pdb_trim
  }
  
  # Plot for the requested protein
  if (show_sse) {
    plotb3(b_factors[[plot_id]], sse = sse_info[[plot_id]],
           typ = "l", ylab = "Bfactor")
  } else {
    plotb3(b_factors[[plot_id]], sse = NULL,
           typ = "l", ylab = "Bfactor")
  }
  
  return(b_factors)
}
```

## Example Run 

```{r}

# This should result in a B-factor plot for our requested structure:
bf <- analyze_protein_bfactors(
  pdb_ids = c("4AKE", "1AKE", "1E4Y"),
  plot_id = "4AKE",
  chain = "A",
  elety = "CA",
  show_sse = TRUE
)

# Showing a small piece of the returned output to confirm the function return:
head(bf[["4AKE"]])

# Plotting another to demonstrate generalizatoin
analyze_protein_bfactors(
  pdb_ids = c("4AKE", "1AKE", "1E4Y"),
  plot_id = "1AKE",
  chain = "A",
  elety = "CA",
  show_sse = FALSE
)
```

This function generalizes the original script by accepting a user-supplied vector of PDB IDs, instead of hard-coding 4AKE, 1AKE, and 1E4Y, and using a loop to apply the same workflow, which is to read the structure, subset to the chosen chain and atom type, and then extract B-factors for any number of input proteins. It then produces the requested B-factor plot for a specific structure chosen via plot_id and returns the B-factor vectors for all inputs, making the analysis reusable and avoiding repeated copy/paste code.
