---
title: "Class 8: Breast Cancer Mini Project"
author: "Matthew Huang (A17978767)"
date: "01/29/2026"
format: pdf
---

## Background

Today we will employ all R techniques for data analysis we used so far including machine learning methods of clustering and PCA to analyze an actual breast cancer biopsy data. 

## Data Import

The data is in CSV format:
```{r}
wiscancer <- read.csv("WisconsinCancer.csv", row.names=1)

head(wiscancer,3)
```
> Q1 How many observations are in this dataset?

```{r}
nrow(wiscancer)
```

> Q2 How many of the observations have a malignant diagnoss?

```{r}
sum(wiscancer$diagnosis == "M")
table(wiscancer$diagnosis)
```

> Q3 How many variables/ features in the data are suffixed with _mean

```{r}
colnames(wiscancer)
length(grep("_mean",colnames(wiscancer)))
```

We need to remove the `diagnosis` column before doing any more analysis of this dataset since we don't want to pass this to PCA etc. Save it as a separate vector that we can later use to compare findings to those of experts. 

```{r}
#saving diagnosis
diagnosis <- wiscancer [,1] 
#dataframe omitting first column
wisc <- wiscancer[,-1]  
```

## Principal Component Analysis (PCA)

The primary function in base R for performing PCA is `prcomp()`. We use the optional argument `scale = TRUE` because the variables in the dataset are measured on different scales.

```{r}
wisc.pr <- prcomp(wisc,scale=T)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.4427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7

## Interpreting PCA Plots

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

It's not a good plot, since not much information is discernable. It is hard to understand due to there just being one large clump of data in the middle, making it too crowded with insufficient scaling to observe correlations visually. 


## Scatter plot observations by components 1 and 2

```{r}
library(ggplot2)
ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
summary(wisc.pr)
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

The PC1 vs PC2 plot shows clearer separation between malignant and benign samples than PC1 vs PC3


```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```

# Calculate variance of components.

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```
# Variance explained by each principal component: pve

```{r}
pve <- pr.var/ sum(pr.var)
```

# Plot variance explained for each principal component

```{r}
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```


# Alternative scree plot of same data, data driven y-axis

```{r}
barplot(pve, ylab = "Explanation For Percent of Variance",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

## ggplot based graph
```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

## Communicating PCA Results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
pc1 <- wisc.pr$rotation [, "PC1"]
wisc.pr$rotation["concave.points_mean",1]
min(pc1)
```

The component of the loading vector for `concave.points_mean` on PC1 is −0.261. This feature has one of the largest absolute loadings on PC1, indicating that it is a major contributor to this principal component.

## Hierarchial clustering 

# Scale the wisc.data data using the "scale()" function
```{r}
wisc.data <- wisc
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist)
plot(wisc.hclust)
abline(h=20, col="pink", lty=2)

```

> Q10. Using plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

~20

## Combining methods

The idea is to use the new variables like the principal component scores stored in `wisc.pr$x` to provide a more concise and informative representation of the dataset than the original features, which were the 30 variables in `wisc.data`, as the basis for clustering analysis.

```{r}
pc.dist <- dist(wisc.pr$x[, 1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

I prefer the ward d2 method, as it looks distributed the best. It is more even, while the single method looks too convoluted, the average method has weird clumps, and the complete method looks way too spread apart. 

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
table(grps, diagnosis)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters, diagnosis)
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```

Now run `table()` with both clustering `grps` and the expert `diagnosis`

```{r}
table(grps, diagnosis)
```
> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

Cluster "1" has 179 "M" Diagnosis
Cluster "2" has 333 "B" Diagnosis

179 TP
24 FP

333 TN
33 FN

Sensitivity: TP / (TP+FN)
```{r}
179/(179+33)
```

Specificity: TN / (TN+FP)
```{r}
333/(333+24)
```

## Prediction

We can use our PCA model for prediction of 

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc

plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="green", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="blue")
```
> Q16 Which of these new patients should we prioritize for follow up based on your results?

Patient 1 should be prioritized for a follow up, as their projected PCA scores place them closer to the malignant cluster.

