---
title: "Class 7: Machine Learning 1"
author: "Matthew Huang (PID: A17978767)"
date: "1/27/26"
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
---

# Background

Today we begin our exploration of important machine learning methods, focusing on **clustering** and **dimensional reduction.**

To start testing these methods, make up some sample data to cluster, where we know what the answer should be.

```{r}
hist( rnorm(3000, mean = 10) )
```

> Q. Can you generate 30 numbers centered at +3 and 30 \@ -3 taken at random from a normal dist?

```{r}
tmp <- c(rnorm(30, mean = 3),
         rnorm(30, mean = -3) )

x <- cbind(x = tmp, y = rev(tmp))
plot(x)

```

## K means clustering

The main function in "base R" for K-means clustering is called `kmeans()`, lets try it

```{r}
k <- kmeans(x, centers = 2)
```

> Q. What component of your kmeans result object has the cluster centers?

```{r}
k$centers
```

> Q. What component of your kmeans result has the cluster size? (how many points in cluster)

```{r}
k$size
```

> Q. What component of your kmeans result has the cluster membership vector? (which points are in cluster)

```{r}
k$cluster
```

> Q. Plot the results of clustering (data colored by the clustering result)

```{r}
plot(x, col = (k$cluster))
points(k$centers, col = "magenta", pch = 15, cex = 1.5)
```

> Q. Run 'kmeans()' again and cluster into 4 groups, plot results like above with coloring by cluster and cluster centers

```{r}
k4 <- kmeans(x, centers = 4)
plot(x, col = k4$cluster)
points(k4$centers, col = "magenta", pch = 15, cex = 1.5)

```

> **Key point:** Kmenas will always return the clustering that we ask for (this is the "K" or "centers" in Kmeans).

```{r}
k$tot.withinss
```

# Hierarchical Clustering

The main function for Hierarchical clustering in base R is called `hclust()`. One of the main differences with respect to the `kmeans()` function is that you can't just pass your input data directly to `hclust()` as it needs a "distance matrix.We can get this from lots of places including the `dist()` function. `hclust()` more flexible.

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

We can "cut the dendrogram or "tree" at a given height to yield our "clusters", for this we use `cutree()`

```{r}
plot(hc)
abline(h = 7, col="red")
grps <- cutree(hc, h = 7)
```

```{r}
grps
```

> Q. Plot our data `x` colored by the clusterinv result from `hclust()`?

```{r}
plot(x, col = grps)
```

# Principal Component Analysis (PCA)

PCA is a popular dimentionality reduction technique widely used in bioinformatics.

## PCA of UK food data

Read data on food consumption in the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

Looks like row names aren't set properly, let's fix this 

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x
```

A better way to do this is to fix the row name assignment at import time. 

```{r}
x <- read.csv(url, row.names = 1)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
x <- dim(x)
x
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I would rather fix the row name assignment at import time, as it appears to be more flexible for usage, and also less complex/convoluted

> Q3. Changing what optional argument in the above barplot() function results in the following plot?

The stacked barplot is produced by setting the optional argument beside = FALSE in the barplot() function. When beside = FALSE, the bars are stacked rather than displayed side by side.

> Q4. Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

The stacked barplot is made by changing the position argument in geom_col() from dodge to stack, since stacking is the default behavior, removing the position argument also results in a stacked barplot.

> Q5.  We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

The points show that if the countries had similar values of the food categories,  the points would lie on the diagonal, if the two countries were being compared have different values for certain food categories then the point will not be on the diagonal.

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

### Heatmap

We can install **pheatmap** package with the `install.packages()` command we used before. Make sure to run in consol and not code chunk in quarto.

```{r}
library(pheatmap)

url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)

pheatmap(as.matrix(x))
library(pheatmap)
pheatmap( as.matrix(x) )
```
Of all these plots ontly the `pairs()` plot was really useful, but this took a bit of work to interpret and will ot scale when I am looking at much bigger databases. 

## PCA the rescue

The main function in "base R" for PCA is called `prcomp()`
```{r}
pca <- prcomp( t(x) )
summary(pca)
```
> Q. How much variance is captured in the first PC?

67.44%

> Q. How many PCs do I need to capture at least 90% of the total variance in the dataset? 

2 PCs, that will capture 96.5% of the total variance

> Q. Plot our main PCA results, people can call this different things depending on their field of study, like PC plot, ordienation plot, score plot, PC1 vs PC2 plot, etc... 

```{r}
attributes(pca)
```
To generate our PCA score plot, we want the `pca$x` component of the result object

```{r}
pca$x
```
```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col=my_cols, pch = 16)

```

```{r}
library(ggplot2)

ggplot(pca$x) + 
  aes(PC1, PC2) +
  geom_point(cols = my_cols)
```







> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}

df <- as.data.frame(pca$x)
df$Country <- rownames(df)

ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()


```

## Digging Deeper: Variable Loadings

How do the original variables (i.e. the 17 differend foods) contribute to our new PCs 

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

> Q9. Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r} 
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```