---
title: "Lab 5: Data Visualization with ggplot"
author: "Matthew Huang (A17978767)"
format: pdf
---

## Background

R has multiple plotting systems. This is a brief comparison of base R graphics and **ggplot2** using the built-in `cars` dataset.

```{r}
head(cars)
```
Base R
```{r}
plot(cars)
```

ggplot2

Install once in the Console: install.packages("ggplot2")
Load each session with library():
```{r}
library(ggplot2)
ggplot(cars)
```

A ggplot typically includes:
	1.	data
	2.	aesthetics (aes())
	3.	one or more geoms
	
```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Speed and Stopping Distance",
    x = "Speed (mph)",
    y = "Stopping distance (ft)"
  ) +
  theme_bw()
```
## Gene Expression Plot

Load gene expression data comparing control and drug-treated samples:
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Simple scatter plot
```{r}
ggplot(genes) +
  aes(x = Condition1, y = Condition2) +
  geom_point(color = "navy", alpha = 0.25, size = 1.5) +
  labs(
    title = "Expression Levels: Control vs Drug",
    x = "Condition 1 (untreated)",
    y = "Condition 2 (treated)"
  ) +
  theme_minimal()
```


Check counts for each expression state:
```{r}
table(genes$State)
```
Color points by expression state:
```{r}
ggplot(genes) +
  aes(x = Condition1, y = Condition2, col = State) +
  geom_point(alpha = 0.6, size = 1.7) +
  scale_color_manual(values = c("purple", "magenta", "pink")) +
  labs(
    title = "Gene Expression Changes After Drug Treatment",
    x = "Control (no drug)",
    y = "Drug treatment"
  ) +
  theme_bw()
```

## Gapminder exploration

Exploring the gapminder dataset:
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
head(gapminder)
```

How many rows are in the dataset?
```{r}
nrow(gapminder)
```

How many continents are represented?
```{r}
table(gapminder$continent)
```

GDP per capita vs life expectancy
```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point(alpha = 0.6, size = 1.7) +
  labs(
    title = "Life Expectancy vs GDP per Capita",
    x = "GDP per capita",
    y = "Life expectancy (years)"
  ) +
  theme_bw()
```

Facet by continent:
```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point(alpha = 0.6, size = 1.7) +
  facet_wrap(~ continent) +
  labs(
    title = "Life Expectancy vs GDP per Capita by Continent",
    x = "GDP per capita",
    y = "Life expectancy (years)"
  ) +
  theme_bw()
```

## First look at dplyr

The dplyr package provides functions for data manipulation, including filter().
```{r}
library(dplyr)
```

Example: Ireland in 2007
```{r}
filter(gapminder, year == 2007, country == "Ireland")
```

Compare two years using faceting:
```{r}
input <- filter(gapminder, year == 2007 | year == 1977)

ggplot(input) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point(alpha = 0.65, size = 1.9) +
  facet_wrap(~ year) +
  labs(
    title = "GDP and Life Expectancy in 1977 vs 2007",
    x = "GDP per capita",
    y = "Life expectancy (years)"
  ) +
  theme_bw()
```